---
title: Mapping Earnings Data from College Scorecard
author: Will Doyle and Benjamin Skinner
output:
  html_document:
    theme: cosmo
---

<!--

PROJ: Analyzing new college score card data
AUTH: Will Doyle and Benjamin Skinner
INIT: 12 September 2015

-->

```{r header, warning = FALSE, message = FALSE, include = FALSE}
################################################################################
## HEADER INFORMATION
################################################################################

## read in strings as strings, not factors
options(stringsAsFactors = FALSE)

##  package list
libs <- c('dplyr',                 # make data wrangling easier
          'gstat',                 # kriging
          'geojsonio',             # to handle geojson data type
          'htmlwidgets',           # way to bind javascript libraries
          'knitr',                 # combine R code and markdown
          'leaflet',               # interactive mapping module
          'maptools',              # mapping projections
          'RColorBrewer',          # create color palettes
          'readr',                 # faster reading of large csv files
          'rgdal',                 # interact with gdal GIS framework
          'rgeos',                 # interact with geos GIS framework               
          'scales',                # Nicely formed scales
          'tidyr')                 # make data wrangling easier

## load packages
lapply(libs, require, character.only = T)

## set relative directory paths
ddir <- '../data/'
mdir <- '../map/'

## projection (do not change unless there's a specific reason)
proj <- CRS('+init=epsg:4269')

## add icon for schools
schoolIcon <- makeIcon(iconUrl = paste0(ddir, 'college-24.svg'),
                       iconWidth = 35, iconHeight = 35)

## set knitr global chunk options
knitr::opts_chunk$set(echo=FALSE,results='hide',warning=FALSE,message=FALSE)

```

```{r externalfunctions}
################################################################################
## READ IN FUNCTIONS FROM EXTERNAL SCRIPT FILE
################################################################################
source('functions.R')
```

```{r options}
################################################################################
## SET YEAR, SCORECARD, AND INSTITUTION VARIABLE OPTIONS
################################################################################

## Set the year for your map. You must have all relevelant files for
## the program to run

## year option (choose any year as long as it's 2011)
year <- 2011

## /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/
## TODO: Three variables: earnigns of entrants,
## county earnings, earnings of entrants as a % of county earnings
## /\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/\/

## outcome column in addition to required
outcome <- 'md_earn_wne_p10'
var_describe <- 'Median earnings of attendees ten years later'

required <- c('unitid','instnm')
sccols <- c(required, outcome)

```

```{r data}
################################################################################
## LOAD DATA
################################################################################

## load county shapefile (national level)
shapefile <- 'gz_2010_us_050_00_500k'

## read in college scorecard data
csdat <- read_csv(paste0(ddir, 'MERGED', year, '_PP.csv'), progress = FALSE)

## set mask based on columns set above
keep <- sapply(sccols, FUN = function(x) { grep(x, tolower(names(csdat))) })

## subset
csdat <- csdat %>% select(keep)

## read in IPEDS institution GEO files
inst_all <- read_csv(paste0(ddir, 'insts.csv'))

## get ACS
acs <- read_csv(paste0(ddir, 'acs_data.csv'))

## get state fips; change names; add leading zeros; drop AK and HI
state_fips <- read_csv(paste0(ddir,'state_fips.csv'))
names(state_fips) <- c('stfips','longname','stabbr')
stfipsvec <- state_fips %>% filter(!stfips %in% c('02','15')) %>% .$stfips

## county layer (same as shapefile name above
county_layer <- shapefile
```

```{r krige}
################################################################################
## MAKE MAP
################################################################################

## merge IPEDS data with CS data
inst_dat <- left_join(inst_all, csdat, by = 'unitid')

## spatial projection of institutions
inst_sp <- instPointsSpatial(dat = inst_dat,
                             proj = proj,
                             outcome_var = outcome
                             )

## REPORT CATEGORIES -----------------------------------------------------------
##
## 1 = Public Doctoral (includes extensive and intensive) 4-year
## 2 = Public Non-Doctoral (all other 4-years)
## 3 = Private Doctoral (as above)
## 4 = Private Non-Doctoral (as above)
## 5 = Public 2 year (>= 90% of degrees at Associate’s or certificate level)
## 6 = Public Vocational/Technical
## 
## -----------------------------------------------------------------------------

## institutional types with report cateogories
insttypes <- list(list('pub2',c(5)))      # public two-years, group = 5


## loop through each type to create layer if not already created
for(i in 1:length(insttypes)) {

    ## NB: if file already exists, don't run; delete or move files to rerun
    if (!file.exists(paste0(mdir,'county_shape_', outcome, '_',
                            insttypes[[i]][[1]], '_', year, '.geojson'))) {

        ## krige if doesn't exist (will take a while)...
        krigeMap(dat = inst_dat,
                 inst_sp = inst_sp,
                 acsdat = acs,
                 select_group = insttypes[[i]][[2]],
                 groupname = insttypes[[i]][[1]],
                 outcome_var = outcome,
                 sf = county_layer,
                 stfipsvec = stfipsvec,
                 ddir = ddir)
        
    } else {
        ## ...else move to the next institutional type
        message(paste0(insttypes[[i]][[1]],' file exists, no new estimation\n'))
        next
    }
}

```

```{r drawmap}
################################################################################
## GENERATE MAPS
################################################################################

## =================================
## PUBLIC TWO-YEAR MAP
## =================================

## read in spatial layer
mapname <- paste0(mdir, 'county_shape_', outcome, '_pub2_', year, '.geojson')   
county_shape <- readOGR(mapname, layer = 'OGRGeoJSON', verbose = FALSE)

pub2_map <- mapSC(countylayer = county_shape,
                  instlayer = inst_sp[inst_sp@data$group == 5,],
                  var_describe = var_describe,
                  outcome_var = outcome)

```

<!--

HTML page content

-->

The map below displays county level estimates of the earnings among working
attendees of local public two-year colleges ten years after first
attendance. The map is interactive: click on any county for county-level
estimate. Zoom using "+/-" in upper left corner. Pan using mouse or
keyboard arrows. Check "institutions" box to see institutional
location. Click on institutions to see scorecard data for that
institution.

## County Level Estimates: Earnings of Public Two-Year Students Ten Years After Entry

```{r printmap_pub2, results = 'show', fig.width = 9.5}
################################################################################
## PRINT MAP
################################################################################

pub2_map
```

Several trends are immediately obvious:

* The "payoff" for college has a strong regional component. Young
people living in the south do not generally live close to any
community colleges that have high attendee incomes.
* Regional variation in earnings is closely tied to proximity to
strong labor markets: institutions in Chicago, New York and the Bay
Area all are in the top deciles of earnings.
* The role of community colleges in economic mobility also stands
out. In many counties, community college attendees earn substantially
more than the median income. These tend (not surprisingly) to be
poorer counties, likely with low overall levels of educational
attainment.


## Notes

The data only include attendees who took some form of federal
financial aid (loans or grants). County-level estimates are based on
the distance-weighted measures for nearby institutions, with nearest
institutions weighted highest. 

## Data

### Median income 10 years after first enrollment

The primary income data come from college scorecard dataset:
[https://collegescorecard.ed.gov/data/](https://collegescorecard.ed.gov/data/).
The variable used is **`r outcome`**. Data are from the year 2011.

### County median income

Median income data come from the American Community Survey 5-year
rolling estimates. The table id is **B22008**. The years surrounding
2011 were used (2008-2012). Data from the table are in the GitHub
respository, but the full table can be
[downloaded here](http://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_12_3YR_S1903&prodType=table).

### Spatial polygons

Spatial shapefiles were taken from the [U.S. Census Bureau Cartographic
Boundary Shapefiles](https://www.census.gov/geo/maps-data/data/cbf/cbf_counties.html).
Specifically, the 500k resolution files were used. These contain fewer
details than finer featured TIGER shapefiles, but are much smaller and
are adequate for this map.

## Download, fork, or collaborate

The GitHub repository for this project may be found at
[https://github.com/wdoyle42/scorecard](https://github.com/wdoyle42/scorecard).

This is an ongoing project. Comments welcome, better code even more so.

## Authors

Will Doyle: [Website](https://my.vanderbilt.edu/willdoyle/),
[GitHub](https://github.com/wdoyle42)  
Benjamin Skinner: [Website](http://btskinner.me/), [GitHub](https://github.com/btskinner)
